Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth

### 2017-10-23 manuscript/manuscript.tex
Edited manuscript

\begin{equation}
    \Delta \rho (r') = | \rho_n(r') - \rho_l(r') |
\end{equation}

@@

\begin{equation}
    \text{max}\{ \Delta \rho(r') \} > \delta
    \label{eq:delta-density}
\end{equation}
\begin{equation}
    \text{max}\{ \Delta \rho(r') \} = \Delta \rho(R_\text{max}),
\end{equation}

@@

\noindent and the value of $b$ depends on the spherical shell on which we performed the test. 
In this way, the shell has a density of 2670 kg/m$^3$ and 3300 kg/m$^3$ on its outer and inner radii, respectively.
The analytical solution of the gravity potential is easily obtained as the sum the solution for a pure exponential function and the one for an homogeneous spherical shell.

@@

Secondly, lets assume the same density function from equation \ref{eq:density-discontinuous}, but now with an asymmetrical distribution: $R_c = (R_1 + R_2)/3$, and perform the same test.
We can see its results in Figure \ref{fig:D-discont-asym}: up to $D=10$ the difference between the numerical model and the analytical solution is greater than 10\%, what
leads us to conclude that the code does not recreates the gravity fields for this case with good accuracy.

@@

As we did with the high varying exponential densities, we can conclude that for arbitrary discontinuous functions a density-based discretization algorithm is also needed in order to guarantee a good performance.

@@

Besides the implementation of this new algorithm, a determination of the maximum value of the delta ratio $\delta$ that produces an accuracy bellow a predefined threshold is also needed in order to guarantee the good performance of the numerical model.

To do this we applied a similar methodology used in the determination of the distance-size ratio $D$: we proposed a set of exponential density functions (as the one in equation \ref{eq:density-exp}) with different values of $b$ and computed the difference between the gravity fields generated by the numerical model with the corresponding analytical solution, each time with a different value of $\delta$ ranging from 0.1 to 1 with a step of 0.1.
This tests have been performed for a thin and a thick spherical shell of 1km and 35km of thickness, respectively.

@@

On Figures \ref{fig:exponential-delta-thin}a and \ref{fig:exponential-delta-thick}a we can see the exponential density functions used on each test for a thin and thick spherical shell, respectively.
While Figures \ref{fig:exponential-delta-thin}b and \ref{fig:exponential-delta-thin}b show their results.

@@

@@

We have developed a Python library that allows us to perform forward gravity computations using tesseroids with variable density in depth.
The code numerically approximates the integral that defines the gravity potential, its gradient and the Marussi tensor through the Gauss-Legendre Quadrature (GLQ).
It essentially consists in approximating the gravity fields of the tesseroids with the effect of point masses located on the scaled nodes of the Legendre polynomials.
We have stated that this method it's well suited to approximate the effect of tesseroids with arbitrary densities, in contrast with the ones that involve Taylor series expansion.

Instead of enhancing the accuracy of the method by increasing the order of the GLQ on each integration, we make use of the modified adaptive discretization algorithm.
It divides the tesseroid by half if the ratio of the distance to the computation point and its size is lower than a predefined distance-size ratio $D$.
We can summarise that $D$ controls the accuracy of the method: a higher value of $D$ generates more divisions of the tesseroids, thus more point masses, what leads to a more precise approximation.

On the other hand, there's no direct relation between the value of $D$ and the error of the computation, that's why we had to determine the minimum value of $D$ that produces an acceptable accuracy for each gravity magnitude.
These distance-size ratio determinations have been made through the comparison of the numerical approximation with the analytical solutions for the exceptional case of a spherical shell.
From the wide range of possible density variations, we've chosen a linear, an exponential and a discontinuous density functions.

The tests over the first two functions have shown that in order to achieve an accuracy of 0.1\%, a value of $D$ equal to 1, 2 and 8 must be used in the calculation of the gravity potential, its gradient components and the components of the Marussi tensor, respectively.
We have also found a limitation on the variation of the exponential density function: the absolute value of the exponent must not be greater than 1.
However, those extremely variations cannot be found inside the Earth, so this limitation does not determines a true restriction for real life models.

On the other hand, the tests over the discontinuous function were not satisfactory.
In such cases we recommend to divide the model in several tesseroids, each one with a specific homogeneous density or a continuous density function, instead of using a single variable density tesseroid.

>>>>

\begin{equation}
    \Delta \rho (r') = | \rho_n(r') - \rho_l(r') |.
\end{equation}

@@

\begin{equation}
    \text{max}\{ \Delta \rho(r') \} > \delta,
    \label{eq:delta-density}
\end{equation}
\begin{equation}
    \text{max}\{ \Delta \rho(r') \} = \Delta \rho(R_\text{max}),
\end{equation}

@@

\noindent and the value of $b$ equal to the thickness of the spherical shell we use to perform the test.
In this way, the density function assumes the values of 2670 kg/m$^3$ and 3300 kg/m$^3$ on the shell's outer and inner radii, respectively.
The analytical solution of the gravity potential is easily obtained as the sum the solution for a pure exponential function and the one for an homogeneous spherical shell.

@@

Secondly, lets assume the same density function from equation \ref{eq:density-discontinuous}, but now with an asymmetrical distribution: $R_c = (R_1 + R_2)/3$, and perform the same test.
We can see its results in Figure \ref{fig:D-discont-asym}: up to $D=10$ the difference between the numerical model and the analytical solution is greater than 10\%, what
leads us to conclude that a density-based discretization algorithm is needed in order to guarantee the accuracy of the model.

@@

@@

Besides the implementation of this new algorithm, we must determine the maximum value of the delta ratio $\delta$ that produces an accuracy bellow a predefined threshold in order to guarantee the good performance of the numerical model.

To do this we applied a similar methodology used in the determination of the distance-size ratio $D$: we proposed a set of exponential density functions (as the one in equation \ref{eq:density-exp}) with different values of $b$, and computed the difference between the gravity fields generated by the numerical model with the corresponding analytical solution, each time with a different value of $\delta$ ranging from 0.1 to 1 with a step of 0.1.
This tests have been performed for a thin and a thick spherical shell of 1km and 35km of thickness, respectively.

@@

On Figures \ref{fig:exponential-delta-thin}a and \ref{fig:exponential-delta-thick}a we can see the exponential density functions used on each test for a thin and thick spherical shell, respectively.
While Figures \ref{fig:exponential-delta-thin}b and \ref{fig:exponential-delta-thin}b show their corresponding results.

@@

Furthermore, we computed the difference between the numerical model and the analytical solution in case of a discontinuous density (as the one in equation \ref{eq:density-discontinuous} with $R_c = (R_1 + R_2)/3$) using a $\delta = 0.2$.
The result for the $g_{zz}$ component of the Marussi tensor is bellow the 0.1\% threshold, achieving a good performance for this default value of $\delta$.

@@

We have developed a Python library that allows us to perform forward gravity computations using tesseroids with variable density in depth.
The code numerically approximates the integral that defines the gravity potential, its gradient and the Marussi tensor through the Gauss-Legendre Quadrature (GLQ).
It essentially consists in approximating the gravity fields of the tesseroids with the effect of point masses located on the scaled nodes of the Legendre polynomials.
We state that this method it's well suited to approximate the effect of tesseroids with arbitrary densities, in contrast with the ones that involve Taylor series expansion.

Instead of enhancing the accuracy of the method by increasing the order of the GLQ on each integration, we make use of the preexisting modified adaptive discretization algorithm and a new density-based discretization algodithm.

The former one divides the tesseroid by half if the ratio of the distance to the computation point and its size is lower than a predefined distance-size ratio $D$.
We can summarise that $D$ controls the accuracy of the method: a higher value of $D$ generates more divisions of the tesseroids, thus more point masses, what leads to a more precise approximation.
Nevertheless it is not sufficient to guarantee the accuracy of the method in case of specific density functions as a highly variable exponential or some discontinuous functions.

For this reason we developed a density-based discretization algorithm that divides the tesseroid on the radius at which the \emph{maximum variation} of the density function takes place, only if a inequation that relates this \emph{maximum variation} with a predefined delta ratio $\delta$ holds.
Therefore the number of subdivisions is indirectly controlled by this delta ratio $\delta$, and thus the accuracy of the computation.

Because there is no direct relation between the values of $D$ and $\delta$ and the error of the computation, we had to empirically determine the minimum value of $D$ and the maximum value of $\delta$ that produce an acceptable accuracy for each gravity field.

The distance-size ratio determinations have been made through the comparison of the numerical approximations with the analytical solutions for the a spherical shell with variable density, which had to be analytically obtained.
From the wide range of possible density variations, we've chosen a linear, an exponential and a discontinuous density functions.
These comparisons have been made without applying the density-based discretization algorithm in order to isolate the other one.
The tests over the first two functions have shown that a value of $D$ equal to 1, 2 and 8 must be used in the calculation of the gravity potential, its gradient components and the components of the Marussi tensor, respectively, in order to achieve an accuracy bellow 0.1\%.
On the other hand we have found a limitation of this algorithm working on its own in case of highly varying exponential functions and in some cases of discontinuous density.

Nevertheless, the density-based discretization algorithm have shown that a value of $\delta = 0.2$ (together with the previously obtained default values of $D$) guarantees a good accuracy on those cases on which the modified adaptive discretization algorithm does not achieve it on its own.
